He-Man: A domain specific language for lightweight concurrency

### 2012-03-11 paper/paper.tex
Graph production.
Wrote abstract and most of introduction; added citations.

% XXX: We need to cite ElliottEtAl somewhere!

%XXX: new title?
\title{He-Man: A domain-specific language for lightweight concurrency}
%\subtitle{I said hey, what's goin' on}
\authorinfo{Carlo Angiuli}{Carnegie Mellon University}{cangiuli@cs.cmu.edu}
\authorinfo{Michael Sullivan}{Carnegie Mellon University}{mjsulliv@cs.cmu.edu}

@@

\begin{abstract}
The event-driven model is the gold standard for highly concurrent network
application programming; while threaded code is easier to write and reason
about, it is considerably less scalable.

We observe that these paradigms are not only interchangeable, but primarily
confined to an application's core control flow; we propose a hybrid model in
which programmers express this core logic at a high level of abstraction, yet
are free to implement the remainder of the application at a low level.

@@

\section{Introduction}
building scalable concurrent network applications. The
language presents a

presents difficult concurrency challenges 
implementation challenge to developers
supporting many simultaneous clients
although memory and bandwidth is plentiful
difficult to design applications
to take full advantage of these resources.
threads inefficient (scheduling)
event loops (hard to write)

\cite{LauerNeedham},

in a natural, straight-line manner while
still producing efficient event loop-based code.

He-Man programs are compact and
measurements indicate that it performs acceptably. Our implementation is
available at
, embedded, domain specific language in Haskell for
building

Network applications need to support many simultaneous clients---this is known
as the C10K (``10,000 client'') problem \cite{Kegel}. Modern servers have enough
memory and bandwidth for massive concurrency, but it is difficult to design
applications to take full advantage of these resources.

A server must maintain separate state for each connection, and often waits for
network or disk I/O to complete before continuing to serve a particular client.
The simplest solution is to maintain one thread per client, using blocking I/O
within each thread. This incurs frequent and expensive context switches at the
operating system level; because the threads have identical code and little
private data, this approach is unnecessary and does not scale.

An improvement is to write multithreaded code in which each thread serves
multiple clients. This is the most popular approach, and is utilized in the
Apache HTTP server \cite{Apache}.

By far the leanest approach is to write an event loop around kernel facilities
for multiplexed, non-blocking I/O, such as \t{select}, \t{poll}, \t{epoll}, or
\t{kqueue}. In this model, all blocking occurs at a single call site in the
code. The programmer is responsible for explicitly maintaining per-client state
and invoking the appropriate client on each I/O event.

This requires the entire application to be restructured as a state machine with
explicit per-client control blocks, a model which is very difficult for
programmers to reason about. There exist numerous C/C++ libraries for event
loops, but none of these avoid the state machine paradigm.

Our observation is that the event loop model is simply trampolining
continuation-passing-style, % TODO citation
which can be obtained by mechanical translation from the straight-line threaded
code. This suggests it is possible to obtain the best of both worlds---a
threaded programming model with the performance of event loops.

To that end, we present He-Man: Haskell Event Manager, Apropos Networking, an
embedded domain-specific language (EDSL) hosted in Haskell, which provides a
threaded abstraction for network application programming and compiles to a C
event loop based around Linux's \t{epoll}.

He-Man is a straightforward monadic language with built-in network primitives,
offering a simple thread-per-client programming model. It is easily extensible
in Haskell, and also allows calls to foreign C functions; it outputs C code
which links with an event loop runtime and any code a programmer might prefer to
write directly in C. Our vision is that C programmers can use He-Man for their
core application logic, and can write the rest of their application in C.

% Contributions; probably should be elsewhere
We identify our contributions as:
\begin{enumerate}

\item We present a case study of compiled embedded domain specific
  language design and implementation.
\end{enumerate}

\section{The event loop model}
%% XXX: it isn't really clear what's going on in this section; it
%% doesn't /really/ talk about event loop stuff in general. It is
%% talking about one possible way to write event based things (the way
%% He-Man generates).

Network applications require a very constrained form of
concurrency. Because in most network applications, I/O (both network
and disk) latency dominates the computation time, each client's thread
of execution spends most of its time waiting on I/O. In particular, we
would like to always yield to a client no longer blocked on I/O; we
need not yield to blocked clients, nor do we (usually) need arbitrary
interleaving of execution.

@@

@@

\pagebreak
\bibliography{citations}{}
\bibliographystyle{abbrvnat}

>>>>

% XXX: We need to cite ElliottEtAl somewhere!
% XXX: standardize on IO vs I/O
% XXX: standardize on domain-specific vs domain specific

%XXX: new title?
\title{He-Man: A domain specific language for lightweight concurrency}
%\subtitle{I said hey, what's goin' on}
\authorinfo{Carlo Angiuli}{Carnegie Mellon University}{cangiuli@cs.cmu.edu}
\authorinfo{Michael Sullivan}{Carnegie Mellon University}{mjsulliv@cs.cmu.edu}

@@

\begin{abstract}
The event-driven model is the gold standard for highly concurrent network
application programming; while threaded code is easier to write and reason
about, it is considerably less scalable. We observe that these paradigms are not
only interchangeable, but primarily confined to an application's core control
flow; we propose a hybrid model in which programmers express this core logic at
a high level of abstraction, yet are free to implement the remainder of the
application at a low level.

@@

% Related work:
%\cite{VanderwaartCrary} re bounded time
%\cite{vonBehrenEtAl} re Capriccio
%\cite{vonBehrenConditBrewer} re cooperative multithreading
%\cite{LiZdancewic}
%\cite{LauerNeedham} also intro
%\cite{MarlowEtAl} also intro

% Functional programming:
%\cite{Vinoski}
%\cite{Hellstrom}
%\cite{BiagioniEtAl}
%\cite{LiEtAl}
%\cite{GraunkeEtAl}
%\cite{Queinnec}

% Haskell EDSLs:
%\cite{Gill}
%\cite{ElliottEtAl}
%\cite{LeijenMeijer}

\section{Introduction}

Modern network servers, like HTTP, IRC, and multiplayer games, need to support
many simultaneous, possibly idle, connections---this is known as the C10K
(``10,000 client'') problem \cite{Kegel}. Because I/O latency is high,
particularly over a network, our goal is to process ready clients while others
are blocked on I/O. To achieve maximum throughput, such servers are usually
written in low-level languages like C.

The simplest way to achieve this concurrency is to use kernel threads, like the
industry-standard Apache web server \cite{Apache}. These incur a fair amount of
overhead, and do not allow control over scheduling; they waste time, for
example, yielding to I/O blocked clients.

More performant web servers, such as Nginx \cite{Nginx} and Lighttpd
\cite{Lighttpd}, use non-blocking I/O routines, multiplexed at a single call
site (the ``event loop''). This allows one thread to serve many clients,
handling each only when its I/O has completed. This event-driven model requires
the routines to explicitly yield to the event loop, forcing programmers to
restructure their code as a state machine, save state across function
boundaries, and implement a scheduler.

Although the event-driven model is typically chosen by developers who wish to
maximize performance (as with Nginx and Lighttpd), von Behren, Condit, and
Brewer \cite{vonBehrenConditBrewer} argue that events are an unnatural
abstraction for this problem domain, and that an appropriate threading library
(in particular, without preemption) can achieve equivalent performance.

In an oft-cited 1979 paper, Lauer and Needham \cite{LauerNeedham} argue that the
threaded and event-driven models are dual, where the thread scheduler plays the
role of an event loop. Then the translation from threads is simply trampolining
continuation-passing-style, where, in this case, I/O operations must yield.

The idea of implementing custom schedulers for lightweight thread abstractions
is not new, and has been used to great effect in functional language runtimes
like GHC \cite{MarlowEtAl}. Li and Zdancewic \cite{LiZdancewic} have developed
monadic abstractions in Haskell for threaded and event-driven concurrency, for
the same problem domain of scalable network applications; their implementation
draws upon the concurrency and synchronization support in the GHC runtime.

We observe that, although concurrency is critical to network applications
because I/O dominates computation time, core I/O routines comprise a small
portion of an application's program logic, which is otherwise devoted to
protocol implementation, caching, etc. We might expect that our concurrency
abstractions can be similarly confined; however, the above solutions require
writing the entire application in Haskell, or at least writing the core in
Haskell and making liberal use of the FFI.

Our contribution is to distill the Lauer-Needham duality to its essence, by
compiling from a simple thread abstraction to self-contained event-driven C
code. Our language, He-Man: Haskell Event Manager, Apropos Networking, is a
compiled domain specific language embedded in Haskell, which offers a small
monadic language supporting thread spawning, blocking events, and external
function calls.

%XXX

The He-Man system consists of a compiler written in Haskell, and a lightweight
runtime in C, which links with the compiler's output. Both are small and easily
extensible. We target C because it is commonly used for network applications,

so it is trivial to link our

any code a programmer might prefer to write directly in C. 
Our vision is that C programmers can use He-Man for their
core application logic, and can write the rest of their application in C.

He-Man programs are compact and measurements indicate that it performs
acceptably.

\section{The event loop model}
Network applications require a very constrained form of
concurrency. Because in most network applications, I/O (both network
and disk) latency dominates the computation time, each client's thread
of execution spends most of its time waiting on I/O. In particular, we
would like to always yield to a client no longer blocked on I/O; we
need not yield to blocked clients, nor do we (usually) need arbitrary
interleaving of execution.

@@

% XXX taken from old intro

There exist C/C++ libraries for event loops\cite{libevent}, but none of these
avoid the state machine paradigm.

@@

% XXX

We identify our contributions as:
* We present the idea of generating code for network facing logic from a
description written in a DSL while building the rest of the application in
C. In a way, the DSL could be considered as a method for programmatically
generating event driven frameworks customized for the application.
* We observe that converting a traditional block-structured imperative
program to a form amenable for use in an event-driven system is analogous
to CPS conversion and thus easily done automatically. (Is this actually a
contribution, or is it obvious and been said a million times before?)
* We present a case study of compiled embedded domain specific language
design and implementation
* We demonstrate that it is practical to build efficient network
applications using our system

\bibliography{citations}{}
\bibliographystyle{abbrvnat}
